---
title: "individual1"
author: "Brandon Rouse"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

```{r}
accident <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable `DAY_WEEK`)?

```{r}
ggplot(accident, aes(x=DAY_WEEK)) + geom_bar() + labs(x = 'Day of Week', y = 'Number of Accidents', title = 'Number of Accidents per Day of the Week')
```

2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable `INJ_SEV`)

```{r}
fatal <- person %>% filter(INJ_SEV == 4)
str(fatal)
```

3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable `MAKE`)

```{r}
```

4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use `readxl::read_xlsx` to read in the GLCs.)

5. Join the accident and person table (work out which variable(s) to use)

```{r}
acc <- person %>% left_join(accident)
```

6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.

```{r}
accMale <- acc %>% mutate(HOUR = as.factor(HOUR)) %>% filter(HOUR != 99) %>% filter(SEX == 1)
ggplot(accMale, aes(x = DAY_WEEK, fill = HOUR)) + geom_bar(position = position_stack()) + labs(x = 'Day of Week', y = 'Number of Accidents', title = 'Number of Accidents involving Males per Hour and Day of the Week')

accFemale <- acc %>% mutate(HOUR = as.factor(HOUR)) %>% filter(HOUR != 99) %>% filter(SEX == 2)
ggplot(accFemale, aes(x = DAY_WEEK, fill = HOUR)) + geom_bar(position = position_stack()) + labs(x = 'Day of Week', y = 'Number of Accidents', title = 'Number of Accidents involving Females per Hour and Day of the Week')

day <- acc %>% group_by(DAY_WEEK) %>% summarize(wrecksPerDay = table(DAY_WEEK))
sort(day$wrecksPerDay, decreasing = TRUE)

hour <- acc %>% group_by(HOUR) %>% summarize(wrecksPerHour = table(HOUR)) %>% filter(wrecksPerHour != 0 & HOUR != 99) 
sort(hour$wrecksPerHour, decreasing = TRUE)

sex <- acc %>% group_by(SEX) %>% summarize(wrecksPerSex = table(SEX)) %>% filter(SEX %in% c(1,2))
sex
```

We can see that Saturday is the day of the week with the most occuring accidents. Saturday is closely followed by Friday and Sunday, meaning that the weekend has a high number of accidents. From the hour table, we can see that most accidents happen at hour '18,' which would be 6 pm. Also, you can see that almost every afternoon hour has more accidents that morning hours. This makes sense since more people are on the road during the afternoon than morning. Finally, you can see that males have almost double the amount of accidents than females. This is contradicting to societal stereotypes, where many say women are worse drivers than men. This proves otherwise. 


7. Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.
```{r}
deaths <- accident %>% left_join(fatal)
deaths %>% ggplot(aes(x=LONGITUD, y = LATITUDE)) + geom_polygon(aes(group=COUNTY)) + geom_point(data = deaths, aes(x=LONGITUD, y = LATITUDE), color = 'lightgreen', size = 0.02, alpha = 0.2) + coord_map(xlim = c(-130,-60), ylim=c(20,50))
```


8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.
```{r}
winter <- deaths %>% filter(WEATHER1 %in% c(1,2,11,12))
summer <- deaths %>% filter(WEATHER1 %in% c(5,6,7,8))
str(summer)
```


28019 accidents in winter
620 in summer
